A Classical Planning Model for Learning, Parsing and Production in Context-Sensitive Languages

### 2018-05-03 doc/aaai-19/CSL.tex
first notes on model recognition

\title{Learning Context-Sensitive Grammars with Classical Planning}
\author{\#39}

@@

\maketitle
\begin{abstract}
This paper presents a novel approach for learning {\em Context-Sensitive Grammars} (CSGs) from small sets of input strings. The approach guarantees that the computed grammar generalizes provided that the input strings are long and diverse enough to cover all the grammar rules.  Our approach is to compile this learning task into a classical planning problem whose solutions are all sequences of actions with a prefix that build the CSG and and a postfix that validates the built grammar in the input strings. The compilation is flexible to implement the three canonical tasks of CSGs, {\it grammar generation}, {\it string production} and {\it string recognition} within the same classical planning model. The experimental validation of our compilation approach reports the empirical data collected when learning the CSGs for the languages $\{a^nb^nc^n : n \geq 1 \}$ and $\{a^{2^n} : n \geq 1 \}$ and shows that overcomes the performance  of a previous compilation for learning {\em Context-Free Grammars}.
\end{abstract}

\section{Introduction}
\label{sec:section1}
A {\em formal grammar} is a set of symbols and production rules that describe how to form the possible strings of certain formal language~\cite{hopcroft:automatatheory:2001}. Usually three canonical tasks are defined over formal grammars:
\begin{itemize}
\item {\it Learning}: Given a set of strings, compute a grammar that is compliant with the input strings.
\item {\it Production}: Given a formal grammar, generate strings that belong to the language represented by the grammar.
\item {\it Recognition} (also known as {\em parsing}): Given a formal grammar and a string, determine whether the string belongs to the language represented by the grammar.
\end{itemize}

Chomsky defined four types of formal grammars that differ in the form and generative capacity of their rules~\cite{chomsky2002syntactic}. Each grammar type generates a different class of formal language that is recognizable with a different kind of automaton: {\bf Type-0} corresponds to the {\em recursively enumerable} languages that can be recognized with a {\em Turing machine}. {\bf Type-1} corresponds to the {\em recursively enumerable} languages that can be recognized with a {\em Linear-bounded Turing machine}. {\bf Type-2} corresponds to {\em context-free} languages that are recognizable with a {\em non-deterministic push-down automaton}. Finally, {\bf type-3} corresponds to {\em regular} languages that can be recognized with a {\em Finite state machine}.

Figure~\ref{fig:csg}(a) shows an example of a CSG grammar ({\em Type-1}) that generates the $\{a^nb^nc^n : n \geq 1 \}$ language. The grammar defines ten production rules, contains three {\em terminal symbols} ($a$, $b$ and $c$) and five {\em non-terminal symbols} ($S$,$B$,$C$,$W$ and $Z$). This CSG can generate, for instance, the string $aaabbbccc$ by applying the sequence of rules $\tup{1,1,0,2,3,4,5,2,3,4,5,2,3,4,5,6,7,7,8,9,9}$. The {\it parse tree} in Figure~\ref{fig:csg}(b) exemplifies this rule application and proves that the string $aaabbbccc$ belongs to the $\{a^nb^nc^n : n \geq 1 \}$ language defined by the CSG grammar.

@@

  \caption{\small (a) Example of a context-sensitive grammar; (b) the corresponding {\it parse tree} for the string $aaabbbccc$.}
  \label{fig:csg}
\end{figure}

Previous work showed that a classical planning compilation can implement the three canonical tasks (namely {\it grammar generation}, {\it string production} and {\it string recognition}) for {\em Type-3} and {\em Type-2} grammars~\cite{segovia2017generating}. In this work we show that a novel classical planning compilation implements these three canonical tasks for grammars of the three last three types ({\em Type-3}, {\em Type-2} and {\em Type-1}). In addition, this novel compilation has fewer input parameters since it does not require to specify any bound on the size of the grammar rules. The experimental validation of our compilation approach reports the empirical data collected when learning the CSGs for the languages $\{a^nb^nc^n : n \geq 1 \}$ and $\{a^{2^n} : n \geq 1 \}$ and shows that overcomes the performance  of a previous compilation for learning {\em Context-Free Grammars}.

@@

\subsection{Context-Sensitive Grammars}
We define a CSGs as a tuple $\mathcal{G}=\tup{V,\Sigma,R}$, where:
\begin{itemize}
\item $V$ is the finite set of non-terminal symbols, also called variables. With $v_0\in V$ is the start non-terminal symbol that represents the whole grammar.
\item $\Sigma$ is the finite set of terminal symbols, which are disjoint from the set of non-terminal symbols, i.e.~$V\cap \Sigma\neq\emptyset$. The set of terminal symbols is the alphabet of the language defined by $\mathcal{G}$
\item Rules in $R$ are defined as $\alpha v\beta\rightarrow \alpha\gamma\beta$ where $v\in V$, $\alpha,\beta\in(V\cup\Sigma)^*$ and $\gamma\in(V\cup\Sigma)^+$. That is, the left-hand and right-hand sides of any production rules $r\in R$ may be surrounded by a context of terminal and nonterminal symbols.
\end{itemize}

@@

For instance, Figure~\ref{fig:csg}(b) shows how the string $S$ yields the string $aaabbbccc$. The language of a CSG, $L(\mathcal{G})=\{e\in \Sigma^*: v_0\Rightarrow^* e\}$, is the set of strings that contain only terminal symbols and that can be yielded from the string that contains only the initial non-terminal symbol $v_0$ (which is denoted by $S$ in the example of Figure~\ref{fig:csg}(b)).

Given a CSG $\mathcal{G}$ and a string $e\in L(\mathcal{G})$ that belongs to its language, we define a {\it parse tree} $t_{\mathcal{G},e}$ as an ordered, rooted tree that determines a concrete syntactic structure of $e$ according to the rules in $\mathcal{G}$:
\begin{itemize}
\item Each node in a parse tree $t_{\mathcal{G},e}$ is either:
\begin{itemize}
\item An {\it internal node} that corresponds to the application of a rule $r\in R$.
\item A {\it leaf node} that corresponds to a terminal symbol $\sigma\in \Sigma$ and has no outgoing branches.
\end{itemize}
\item Edges in a parse tree $t_{\mathcal{G},e}$ connect non-terminal symbols to terminal or non-terminal symbols following the rules $R$ in $\mathcal{G}$.
\end{itemize}

\subsection{Linear-bounded Turing Machines}
We define a {\em Turing machine} as a tuple $M=\tup{Q,q_o,Q_{\bot},\mathcal{T},\square,\Sigma,\delta}$, where:
\begin{itemize}
\item $Q$, is a finite and non-empty set of machine states such that $q_0\in Q$ it the initial state of the machine and $Q_{\bot}\subseteq Q$ is the subset of terminal states.  
\item $\mathcal{T}$ is the {\em tape alphabet}, that is a finite an non-empty set of symbols that includes the {\em blank symbol} $\square\in\mathcal{T}$ (the only symbol allowed to occur on the tape infinitely often at any step during the computation).
\item $\Sigma$ is the {\em input alphabet}, the set of symbols allowed to initially appear in the tape.
\item $\delta: (Q\setminus Q_{\bot})\times \mathcal{T}\rightarrow Q\times \mathcal{T}\times\{left,right\}$ is the {\em transition function}. If $\delta$ is not defined for the current state of the machine and the current tape symbol, then the machine halts.
\end{itemize}

A table is the most common convention to represent the transitions defined by $\delta$, where the table rows are indexed with the current tape symbol, while the table columns are indexed by the current machine state. For each possible tape symbol and state of the machine there is a table entry that defines: (1) the tape symbol to print at the current position of the header (2) whether the header is shifted {\em left} or {\em right} after the print operation and (3), the new state of the machine after the print operation. For instance, Figure~\ref{tab:tm-anbncn} shows the table that represents the $\delta$ function of a {\em Turing Machine} for recognizing the $\{a^nb^nc^n : n \geq 1 \}$ language. In this example the tape alphabet is $\Sigma=\{a,b,c,x,y,z,\square\}$ while the possible machine states are $Q=\{q_0,q_1,q_2,q_3,q_4,\underline{q_5}\}$ where \underline{$q_5$} is the acceptor state.

\begin{figure}
\begin{center}
    \begin{tabular}{| c | c | c | c | c | c | c |}
    \hline
      & $q_0$ & $q_1$ & $q_2$ & $q_3$ & $q_4$ & \underline{$q_5$} \\ \hline
    a & x,r,$q_1$ & a,r,$q_1$ & - &  a,l,$q_3$ & - & - \\ \hline
    b & - & y,r,$q_2$ & b,r,$q_2$ & b,l,$q_3$ & - & - \\ \hline
    c & - & - & z,l,$q_3$ & - & - & - \\ \hline
    x & - & - & - & x,r,$q_0$ & - & - \\ \hline
    y & y,r,$q_4$ & y,r,$q_1$ & - & y,l,$q_3$ & y,r,$q_4$ & - \\ \hline
    z & - & - & z,r,$q_2$ & z,l,$q_3$ & z,r,$q_4$ & - \\\hline
    $\square$ & - & - & - & - & $\square$,r,$q_5$  & - \\                
    \hline
    \end{tabular}
\end{center}
  \caption{\small Example of a seven-symbol six-state {\em Turing Machine} for recognizing the $\{a^nb^nc^n : n \geq 1 \}$ language (\underline{$q_5$} is the acceptor state).}
  \label{tab:tm-anbncn}
\end{figure}

A {\em Turing machine} is an abstract model of computation that operates on an infinite memory tape. An approach to implement in practice a Turing machine, is to bound the size of its tape. This can be done by: 
\begin{enumerate}
\item Extending the input alphabet with two special symbols, serving as {\em left} and {\em right endmarkers}.
\item Limiting computation to the portion of the tape containing the input plus the two tape cells holding the endmarkers. This means that transitions may not print other symbols over the endmarkers and that may neither move to the left of the left endmarker nor to the right of the right endmarker.
\end{enumerate}

@@

Our approach to learning CSGs is to leverage on classical planning to learn an \strips\ action model s.t. each action schema in the model encodes a transition of a {\em Linear-bounded Turing Machine} that recognices the strings given as input.

First we show how an entry in the table that defines the transitions of a {\em Linear-bounded Turing Machine} can be modeled as \strips\ action schema. Next we show how classical planning can be effectively used to learn action models of this kind.

\subsection{Modeling a Linear-bounded Turing Machine with Classical Planning}
Given $\delta$, the {\em transition function} of a {\em Linear-bounded Turing Machines} $M$, it can be encoded as a classical planning frame $\Phi=\tup{F,A}$ as follows.

We assume that fluents $F$ are instantiated from a set of {\em predicates} $\Psi$, as in PDDL~\cite{fox2003pddl2}. Each predicate $p\in\Psi$ has an argument list of arity $ar(p)$. Given a set of {\em objects} $\Omega$ that represent the cells in the tape of the given Turing Machine $M$, the set of fluents $F$ is induced by assigning objects in $\Omega$ to the arguments of the predicates in $\Psi$; i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$, where $\Omega^k$ is the $k$-th Cartesian power of $\Omega$. In more detail there are four predicates in $\Psi$:
\begin{itemize}
\item {\tt (head ?x)} that encodes the current position of the header in the tape.
\item {\tt (next ?x1 ?x2)} that encodes that the cell {\tt ?x2} follows cell {\tt ?x1} in the tape.
\item {\tt (symbol-$\sigma$ ?x)} that encodes that the tape cell {\tt ?x} contains the symbol $\sigma\in\Sigma$.
\item {\tt (state-$q$)} that encodes that $q\in Q$ is the current machine state.
\end{itemize}

Likewise we assume that actions $a\in A$ are instantiated from \strips\ operator schema. For each transition in $\delta$, a \strips\ action model is defined such that:
\begin{itemize}
\item The {\bf header} of the schema is {\tt rule-id(?xl ?x ?xr)} where $id$ uniquely identifies the transition in $\delta$ and the parameters $?xl$, $?x$ and $?xr$ are tape cells.
\item The {\bf precoditions} of the schema includes {\tt(head ?x)} and {\tt (next ?xl ?x) (next ?x ?xr)} to force that $?x$ is the current tape cell pointed by the header and that $?xl$ and $?xr$ respectively are its left and right neighbours. Additionally the schema includes preconditions {\tt(symbol-$\sigma$ ?x)} and {\tt (state-$q$)} to capture the current symbol pointed by the header and the currrent machine state.
\item The {\bf delete effects} remove the current symbol pointed by the header and the currrent machine state. Finally, the {\bf positive effects} set the new symbol pointed by the header and the new machine state 
\end{itemize}

To illustrate this, Figure~\ref{fig:update-rule} shows the rule $a,q_0\rightarrow x,r,q_1$ of the Turing Machine defined in Figure~\ref{tab:tm-anbncn}. The full encoding of the Turing Machine defined in Figure~\ref{tab:tm-anbncn} produces a total of sixteen \strips\ action schema with the same structure as the one of Figure~\ref{fig:update-rule}. 
\begin{figure}[hbt!]
\begin{scriptsize}
\begin{verbatim}
(:action rule1 ;; a,q_0 -> x,r,q_1
  :parameters (?xl ?x ?xr)
  :precondition (and (head ?x) 
                     (next ?xl ?x) (next ?x ?xr) 
                     (symbol-a ?x) (state-q_0))
  :effect (and (not (head ?x)) 
               (not (symbol-a ?x)) 
               (not (state-q_0))
               (symbol-x ?x) (head ?xr) (state-q_1)))
\end{verbatim}
\end{scriptsize}
 \caption{\small \strips\ action schema that models the first rule $a,q_0\rightarrow x,r,q_1$ of the Turing Machine defined in Figure~\ref{tab:tm-anbncn}.}
\label{fig:update-rule}
\end{figure}

With this regard, the execution of a {\em Linear-bounded Turing Machine} can then be defined as a {\em plan trace} $\mathcal{T}=\tup{s_0,a_1,s_1,\ldots,a_n,s_n}$ such that $s_0$ encodes the initial state of the tape as well as the initial machine state and for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$.

\subsection{Learning \strips\ Models with Classical Planning}
Here we show that the existing classical planning compilation for learning \strips\ action schemes from plan traces~\cite{aineto:learningSTRIPS:ICAPS2018} can be adapted to address the learning of CSGs. Note that in this case both the preconditions and negative effects of the \strips\ action schemes are already known so the learning task reduces to learning positive effects for these same action schemes.

We formalize the learning of \strips\ action schemes from plan traces as a tuple $\Lambda=\tup{\mathcal{M},\mathcal{T},\Psi}$ where:
\begin{itemize}
\item $\mathcal{M}$, the set of {\em partially specified} operator schemas.
\item $\mathcal{T}=\tup{s_0,a_1,s_1,\ldots,a_n,s_{n}}$ is a {\em plan trace} obtained watching the execution of the classical plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. 
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning frame. 
\end{itemize}

@@

Given a learning task $\Lambda=\tup{\mathcal{M},\mathcal{T},\Psi}$ the compilation defined by~\citeauthor{aineto:learningSTRIPS:ICAPS2018} outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ built instantiating the predicates $\Psi$ with the objects $\Omega$ that appear in the plan trace. Formally, $\Omega=\bigcup_{\small s\in\mathcal{O}} obj(s)$, where $obj$ is a function that returns the objects that appear in a given state.
\item Fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$, that represent the programmed action model. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative/positive effect in the schema $\xi\in \mathcal{M}'$. 
\item The fluents $mode_{prog}$ and $mode_{val}$ to indicate whether the operator schemas are programmed or validated, and the fluents $\{test_i\}_{1\leq i\leq n}$, indicating the state where the learned action model is validated.
\end{itemize}
\item $I_{\Lambda}$ encodes the first observation, $s_0\subseteq F$, the fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ that encode the {\em partially specified} operator schemas and sets $mode_{prog}$ to true. 
\item $G_{\Lambda}=\bigcup_{1\leq i\leq n}\{test_i\}$, requires that the programmed action model is validated in all de states of the input trace.
\item $A_{\Lambda}$ comprises three kinds of actions:
\begin{enumerate}
\item Actions for {\em programming} operator schema $\xi\in\mathcal{M}$. For our purpose we only need to define the actions that add a {\em positive} effect $f\in F_v(\xi)$ to the action schema $\xi\in\mathcal{M}$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programEff_{f,\xi}})=&\{\neg del_{f}(\xi),\neg add_{f}(\xi), mode_{prog}\},\\
\cond(\mathsf{programEff_{f,\xi}})=&\{pre_{f}(\xi)\}\rhd\{del_{f}(\xi)\},\\
& \{\neg pre_{f}(\xi)\}\rhd\{add_{f}(\xi)\}.
\end{align*}
\end{small}

>>>>

\title{A Classical Planning Model for Learning, Parsing and Production in Context-Sensitive Languages}
\author{\#39}

@@

\maketitle
\begin{abstract} 
This paper presents a novel approach for learning the acceptor automaton of a given {\em Context-Sensitive Language} (CSLs) from small sets of input strings. Our approach is to compile this learning task into a classical planning problem whose solutions are sequences of actions s.t. their prefix builds the acceptor automaton and their postfix validates that the built automaton accepts the input strings. The compilation is flexible to implement the three canonical tasks of Context-Sensitive Grammars (CSGs), {\it learning}, {\it parsing} and {\it production} within the same classical planning model. The paper reports the empirical data collected when learning the acceptor automata for the CSLs $\{a^nb^nc^n : n \geq 1 \}$ and $\{a^{2^n} : n \geq 1 \}$ and shows that our compilation outperforms a previous classical planning compilation for  learning {\em Context-Free Grammars}.
\end{abstract}

\section{Introduction}
\label{sec:section1}
A {\em formal grammar} is a set of symbols and production rules that describe how to form the possible strings of certain formal language~\cite{hopcroft:automatatheory:2001}. Usually three canonical tasks are defined over formal grammars:
\begin{itemize}
\item {\it Learning}: Given a set of strings, compute a grammar that is compliant with the input strings.
\item {\it Recognition} (also known as {\em parsing}): Given a formal grammar and a string, determine whether the string belongs to the language represented by the grammar.
\item {\it Production}: Given a formal grammar, generate a string that belongs to the language represented by the grammar.
\end{itemize}

Chomsky defined four types of formal grammars that differ in the form and generative capacity of their rules~\cite{chomsky2002syntactic}. Each grammar type generates a different class of formal language that is recognizable with a different kind of acceptor automaton: {\bf Type-0} corresponds to the {\em recursively enumerable} languages that can be recognized with a {\em Turing machine}. {\bf Type-1} corresponds to the {\em recursively enumerable} languages that can be recognized with a {\em linear-bounded Turing machine}. {\bf Type-2} corresponds to {\em context-free} languages that are recognizable with a {\em non-deterministic push-down automaton}. Finally, {\bf type-3} corresponds to {\em regular} languages that can be recognized with a {\em finite state machine}.

Figure~\ref{fig:csg}(a) shows an example of a CSG grammar ({\em Type-1}) that generates the $\{a^nb^nc^n : n \geq 1 \}$ language. The grammar defines ten production rules over three {\em terminal symbols} ($a$, $b$ and $c$) and five {\em non-terminal symbols} ($S$,$B$,$C$,$W$ and $Z$). This CSG produces the string $aaabbbccc$ applying the sequence of rules $\tup{1,1,0,2,3,4,5,2,3,4,5,2,3,4,5,6,7,7,8,9,9}$. The {\it parse tree} in Figure~\ref{fig:csg}(b) exemplifies this rule application and proves that the string $aaabbbccc$ belongs to the $\{a^nb^nc^n : n \geq 1 \}$ language.

@@

  \caption{\small (a) Example of a {\em context-sensitive grammar}; (b) the corresponding {\it parse tree} for that grammar and the string $aaabbbccc$.}
  \label{fig:csg}
\end{figure}

Previous work showed that a classical planning compilation can implement {\it learning}, {\it parsing} and {\it production} for {\em Type-3} and {\em Type-2} languages~\cite{segovia2017generating}. In this work we show that a novel classical planning compilation implements these three canonical tasks for {\em Type-3}, {\em Type-2} and {\em Type-1} languages. In addition, this novel compilation has fewer input parameters since it does not require to bound the size of the grammar rules. The evaluation of our compilation approach reports the empirical data collected when learning the acceptor automata for the CSLs $\{a^nb^nc^n : n \geq 1 \}$ and $\{a^{2^n} : n \geq 1 \}$ and shows that it outperforms of a previous compilation for {\it learning}, {\it parsing} and {\it production} in {\em Context-Free Languages}.

@@

\subsection{Context-Sensitive Grammars}
We define a CSGs as a tuple $\mathcal{G}=\tup{V,\Sigma,R}$, where:
\begin{itemize}
\item $V$ is the finite set of {\em non-terminal symbols}, also called variables. With $v_0\in V$ is the start non-terminal symbol that represents the whole grammar.
\item $\Sigma$ is the finite set of {\em terminal symbols}, which are disjoint from the set of non-terminal symbols, i.e.~$V\cap \Sigma\neq\emptyset$. The set of terminal symbols is the {\em alphabet} of the language defined by $\mathcal{G}$
\item $R$ is a finite set of $\alpha v\beta\rightarrow \alpha\gamma\beta$ rules where $v\in V$, $\alpha,\beta\in(V\cup\Sigma)^*$ and $\gamma\in(V\cup\Sigma)^+$. That is, the left-hand and right-hand sides of any production rule $r\in R$ may be surrounded by a context of terminal and non-terminal symbols.
\end{itemize}

@@

Figure~\ref{fig:csg}(b) shows how the string $S$ yields the string $aaabbbccc$. The language of a CSG, $L(\mathcal{G})=\{e\in \Sigma^*: v_0\Rightarrow^* e\}$, is the set of strings that contain only terminal symbols and that can be yielded from the string that contains only the initial non-terminal symbol $v_0$ (denoted by $S$ in the example of Figure~\ref{fig:csg}(b)).

Given a CSG $\mathcal{G}$ and a string $e\in L(\mathcal{G})$ that belongs to its language, we define a {\it parse tree} $t_{\mathcal{G},e}$ as an ordered, rooted tree that determines a concrete syntactic structure of $e$ according to the rules in $\mathcal{G}$:
\begin{itemize}
\item Each node in a parse tree $t_{\mathcal{G},e}$ is either: An {\it internal node} that corresponds to the application of a rule $r\in R$ or a {\it leaf node} that corresponds to a terminal symbol $\sigma\in \Sigma$.
\item Edges in a parse tree $t_{\mathcal{G},e}$ connect non-terminal symbols to terminal or non-terminal symbols following a rule in $\mathcal{G}$.
\end{itemize}

\subsection{Linear-bounded Turing Machines}
A {\em Turing machine} is a tuple $M=\tup{Q,q_o,Q_{\bot},\mathcal{T},\square,\Sigma,\delta}$:
\begin{itemize}
\item $Q$, is a finite and non-empty set of machine states such that $q_0\in Q$ is the initial state of the machine and $Q_{\bot}\subseteq Q$ is the subset of acceptor states.  
\item $\mathcal{T}$ is the {\em tape alphabet}, that is a finite non-empty set of symbols that includes the {\em blank symbol} $\square\in\mathcal{T}$ (the only symbol allowed to occur on the tape infinitely often) and that contains $\Sigma\subseteq\mathcal{T}$, the set of symbols allowed to initially appear in the tape (also called the {\em input alphabet}).
\item $\delta: (Q\setminus Q_{\bot})\times \mathcal{T}\rightarrow Q\times\{left,right\}\times \mathcal{T}$ is the {\em transition function}. If $\delta$ is not defined for the current pair of machine state and tape symbol, then the machine halts.
\end{itemize}

A table is the most common convention to represent the transitions defined by $\delta$, where the table rows are indexed by the current tape symbol, while the table columns are indexed by the current machine state. For each possible pair of tape symbol and machine state, there is a table entry that defines: (1) the tape symbol to print at the current position of the header (2) whether the header is shifted {\em left} or {\em right} after the print operation and (3), the new state of the machine after the print operation. For instance, Figure~\ref{tab:tm-anbncn} shows the table that represents the $\delta$ function of a {\em Turing Machine} for recognizing the $\{a^nb^nc^n : n \geq 1 \}$ language. In this example the tape alphabet is $\Sigma=\{a,b,c,x,y,z,\square\}$ while the possible machine states are $Q=\{q_0,q_1,q_2,q_3,q_4,\underline{q_5}\}$ where \underline{$q_5$} is the only acceptor state.

\begin{figure}
\begin{center}
    \begin{tabular}{| c | c | c | c | c | c | c |}
    \hline
      & $q_0$ & $q_1$ & $q_2$ & $q_3$ & $q_4$ & \underline{$q_5$} \\ \hline
    a & x,r,$q_1$ & a,r,$q_1$ & - &  a,l,$q_3$ & - & - \\ \hline
    b & - & y,r,$q_2$ & b,r,$q_2$ & b,l,$q_3$ & - & - \\ \hline
    c & - & - & z,l,$q_3$ & - & - & - \\ \hline
    x & - & - & - & x,r,$q_0$ & - & - \\ \hline
    y & y,r,$q_4$ & y,r,$q_1$ & - & y,l,$q_3$ & y,r,$q_4$ & - \\ \hline
    z & - & - & z,r,$q_2$ & z,l,$q_3$ & z,r,$q_4$ & - \\\hline
    $\square$ & - & - & - & - & $\square$,r,$q_5$  & - \\                
    \hline
    \end{tabular}
\end{center}
  \caption{\small Example of a seven-symbol six-state {\em Turing Machine} for recognizing the $\{a^nb^nc^n : n \geq 1 \}$ language (\underline{$q_5$} is the only acceptor state).}
  \label{tab:tm-anbncn}
\end{figure}

A {\em Turing machine} is an abstract model of computation that operates on an infinite memory tape. An approach to implement in practice a Turing machine, is to bound the size of its tape. This can be done by: 
\begin{enumerate}
\item Extending the input alphabet with two special symbols that serve as the {\em left} and {\em right endmarkers}.
\item Limiting computation to the portion of the tape that contains the input plus the tape cells holding between the two endmarkers. This means that transitions do not print other symbols over the endmarkers and that never move neither to the left of the left endmarker nor to the right of the right endmarker.
\end{enumerate}

@@

Our approach is to leverage classical planning to learn an \strips\ action model s.t. each action schema in the model encodes a transition of the aimed {\em Linear-bounded Turing Machine}. First we show how an entry in the table that defines the transitions of a {\em Linear-bounded Turing Machine} is modeled as a \strips\ action schema. Next we show how classical planning can be effectively used to learn action models of this kind.

\subsection{\strips\ Modeling of a Linear-bounded Turing Machine}
A classical planning frame $\Phi=\tup{F,A}$ encodes the {\em transition function} $\delta$ of a {\em Linear-bounded Turing Machine} $M$ as follows.

We assume that fluents $F$ are instantiated from a set of {\em predicates} $\Psi$, as in PDDL~\cite{fox2003pddl2}. Each predicate $p\in\Psi$ has an argument list of arity $ar(p)$. Given a set of {\em objects} $\Omega$ that represent the cells in the tape of the given Turing Machine $M$, the set of fluents $F$ is induced by assigning objects in $\Omega$ to the arguments of the predicates in $\Psi$; i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$, where $\Omega^k$ is the $k$-th Cartesian power of $\Omega$.

The predicates in $\Psi$ are:
\begin{itemize}
\item {\tt (head ?x)} that encodes the current position of the header in the tape.
\item {\tt (next ?x1 ?x2)} encoding that the cell {\tt ?x2} follows cell {\tt ?x1} in the tape.
\item {\tt (symbol-$\sigma$ ?x)} encoding that the tape cell {\tt ?x} contains the symbol $\sigma\in\Sigma$.
\item {\tt (state-$q$)} encoding that $q\in Q$ is the current machine state.
\end{itemize}

Likewise we assume that actions $a\in A$ are instantiated from \strips\ operator schema. For each transition in $\delta$, a \strips\ action schema is defined such that:
\begin{itemize}
\item The {\bf header} of the schema is {\tt transition-id(?xl ?x ?xr)} where $id$ uniquely identifies the transition in $\delta$ and the parameters $?xl$, $?x$ and $?xr$ are tape cells.
\item The {\bf precoditions} of the schema includes {\tt(head ?x)} and {\tt (next ?xl ?x) (next ?x ?xr)} to force that $?x$ is the tape cell currently pointed by the header and that $?xl$ and $?xr$ respectively are its left and right neighbours. Additionally the schema includes preconditions {\tt(symbol-$\sigma$ ?x)} and {\tt (state-$q$)} to capture the symbol pointed by the header and the currrent machine state.
\item The {\bf delete effects} remove the symbol pointed by the header and the currrent machine state while the {\bf positive effects} set the new symbol pointed by the header and the new machine state.
\end{itemize}

The \strips\ action schema of Figure~\ref{fig:update-rule} models the rule $a,q_0\rightarrow x,r,q_1$ of the Turing Machine defined in Figure~\ref{tab:tm-anbncn}. The full encoding of the Turing Machine defined in Figure~\ref{tab:tm-anbncn} produces a total of sixteen \strips\ action schema with the same structure as the one of Figure~\ref{fig:update-rule}. 
\begin{figure}[hbt!]
\begin{scriptsize}
\begin{lstlisting}
(:action transition-1 ;; a,$q_0\rightarrow$ x,r,$q_1$
  :parameters (?xl ?x ?xr)
  :precondition (and (head ?x)                       
                     (symbol-a ?x) (state-$q_0$)
                     (next ?xl ?x) (next ?x ?xr))
  :effect (and (not (head ?x)) 
               (not (symbol-a ?x)) (not (state-$q_0$))
               (head ?xr) (symbol-x ?x) (state-$q_1$)))
\end{lstlisting}
\end{scriptsize}
 \caption{\small \strips\ action schema that models the transition $a,q_0\rightarrow x,r,q_1$ of the Turing Machine defined in Figure~\ref{tab:tm-anbncn}.}
\label{fig:update-rule}
\end{figure}

The execution of a {\em Linear-bounded Turing Machine} can then be defined as a {\em plan trace} $\mathcal{T}=\tup{s_0,a_1,s_1,\ldots,a_n,s_n}$ such that $s_0$ encodes the initial state of the tape plus the initial machine state and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$.

\subsection{Learning \strips\ Models with Classical Planning}
Learning \strips\ action schemes from plan traces is formalied as a $\Lambda=\tup{\mathcal{M},\mathcal{T},\Psi}$ tuple, where:
\begin{itemize}
\item $\mathcal{M}$ is the set of {\em partially specified} operator schemas.
\item $\mathcal{T}=\tup{s_0,a_1,s_1,\ldots,a_n,s_{n}}$ is a {\em plan trace} obtained watching the execution of the classical plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. 
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning frame. 
\end{itemize}

@@

Given a learning task $\Lambda=\tup{\mathcal{M},\mathcal{T},\Psi}$ the compilation defined by~\citeauthor{aineto:learningSTRIPS:ICAPS2018} outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$. A solution plan to the classical planning task $P_{\Lambda}$ resulting from this compilation is a sequence of actions that:
\begin{enumerate}
\item Programs the action model $\mathcal{M}'$. A solution plan starts with a {\em prefix} that, for each action schema $\xi\in\mathcal{M}$, determines its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item Validates the action model $\mathcal{M}'$. The solution plan continues with a postfix that reproduces the given plan trace with the programmed action model $\mathcal{M}'$.
\end{enumerate}

\subsection{Learning CSGs with Classical Planning}
Learning the acceptor automaton of a CSL is a particular case of the $\Lambda=\tup{\mathcal{M},\mathcal{T},\Psi}$ learning task such that:
\begin{itemize}
\item $\mathcal{M}$ is the set of {\em partially specified} operator schemas that encodes the preconditions and negative effects of the \strips\ schemas that encode the {\em transition function} $\delta$ of the {\em Linear-bounded Turing Machine} $M$.
\item $\mathcal{T}$ represents the execution of the {\em Linear-bounded Turing Machine} $M$ for the parsing of a given string.
\item $\Psi=\{${\small\tt(head ?x),(next ?x1 ?x2),(symbol-$\sigma$ ?x),(state-q)}$\}$ . 
\end{itemize}

Here we show that the existing classical planning compilation $P_{\Lambda}$ for learning \strips\ action schemes from plan traces~\cite{aineto:learningSTRIPS:ICAPS2018} can be adapted to learning the acceptor automaton of a given CSL. When learning a CSGs the input {\em plan trace} $\mathcal{T}$ is partially observed:
\begin{enumerate}
\item States $s_i$, with $1\leq i\leq n$, do not contain the fluents {\tt (state-q)} s.t. $q\in Q$.
\item At each state, $s_i$ with $1\leq i\leq n$, the exact executed action $a_i$ is unobserved while it is known that $a_i$ is built instantiaing the schema corresponding to a transition rule $\sigma,q\rightarrow \sigma',\{r,l\},q'$ provided that $\sigma$ is the symbol at the tape cell pointed by the header at state $s_i$.
\end{enumerate}
To handle the first source of partial observability the compilation does not require any modification because $s_0$ is fully observable since initially the machine state is always $q_0$ (likewise the machine state is always $q_n$ at the last state that corresponds to a string accepted by the learned automaton) and the compilation is robust to missing fluents in the intermediate states of the input plan trace $\mathcal{T}$.

Addressing the second source of partial observability requires a modification of the $P_{\Lambda}$ compilation. Adding extra preconditions to the apply actions so they are  only appliable actions iff they are built instantiaing the schema corresponding to the transition $\sigma,q\rightarrow \sigma',{r,l},q'$ provided that $\sigma$ is the symbol at the cell tape that is currently pointed by the header at state $s_i$.
